---
title: "Getting started with cumba"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cumba}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE
)
```

# What is `cumba`?

**CUMBA** (Carbon Use Model for yield and Brix Assessment) is a daily
time-step crop model for **processing tomato**. Given daily weather and
an irrigation schedule, it computes:

- thermal time, phenology and root growth,
- a soil-water balance on three layers,
- carbon accumulation modulated by heat, cold and water stress,
- fresh-fruit yield and **Brix** at harvest.

The package exposes two top-level entry points:

- `cumba_experiment()` — run the model on one or more **observed
  experiments**, using a user-supplied `irrigation_df`.
- `cumba_scenario()`   — run the model in **deficit-irrigation scenario
  mode**, where irrigation is automatically triggered when the simulated
  water-stress falls below a phase-specific threshold.

```{r setup}
library(cumba)
```

# Default parameters

A complete default parameter set is bundled with the package:

```{r params}
str(cumbaParameters, max.level = 1)
cumbaParameters$Tbase
```

`cumbaParameters` is a *named list of lists* with fields
`value`, `min`, `max`, `description`. Both `cumba_experiment()` and
`cumba_scenario()` accept this format directly and convert it
internally into a tibble of values.

# A bundled example dataset: `tomatoFoggia`

The package ships with a real-world dataset from processing-tomato
field trials in Foggia (Apulia, Italy):

```{r data}
data(tomatoFoggia)
names(tomatoFoggia)
head(tomatoFoggia$weather)
head(tomatoFoggia$irrigation)
head(tomatoFoggia$production)
```

# A first run

The minimal call uses the default parameters and the bundled weather
and irrigation data:

```{r run, eval = FALSE}
weather       <- tomatoFoggia$weather
irrigation_df <- tomatoFoggia$irrigation

out <- cumba_experiment(
  weather       = weather,
  param         = cumbaParameters,
  irrigation_df = irrigation_df,
  estimateRad   = TRUE,
  estimateET0   = TRUE,
  fullOut       = FALSE
)

head(out)
```

The output is a `data.frame` with one row per (site, year, experiment,
day-of-year). With `fullOut = FALSE` you get the *short* output:

| column        | meaning                          |
|---------------|----------------------------------|
| `site`        | site name                        |
| `year`        | calendar year                    |
| `experiment`  | experiment id                    |
| `doy`         | day of year                      |
| `p`           | precipitation (mm)               |
| `irrigation`  | irrigation (mm)                  |
| `stage`       | phenological stage               |
| `swc`         | soil water content (%)           |
| `yield`       | fresh fruit yield (q ha-1)       |
| `brix`        | actual Brix (°)                  |

Set `fullOut = TRUE` to get every internal state variable (carbon,
flowering, fruit-water content, Kc, stresses, etc.).

# What's next?

- See **Run an experiment** (`vignette("cumba-experiment")`) for a
  complete experiment-mode walkthrough on `tomatoFoggia`.
- See **Run a scenario** (`vignette("cumba-scenario")`) for the
  deficit-irrigation mode.
- See **Calibrate ET₀ and radiation**
  (`vignette("estimate_et0_rad")`) for the radiation/ET₀ workflow.
