---
title: "Run a deficit-irrigation scenario with cumba_scenario()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run a deficit-irrigation scenario with cumba_scenario()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE,
  eval    = FALSE
)
```

# Why `cumba_scenario()`?

While `cumba_experiment()` reproduces an observed irrigation schedule,
`cumba_scenario()` runs the model in **automatic deficit-irrigation
mode**. Irrigation is triggered when, simultaneously:

1. the simulated water-stress index `ws` falls **below a phase-specific
   threshold** (`waterStressLevel`), and
2. at least `minimumTurn` days have elapsed since the previous wetting.

The amount of water applied refills the soil to **field capacity** in
the rooted layers.

This is the function to use for *what-if* studies, irrigation strategy
design, and climate-impact assessments.

# Phase-specific stress thresholds and turns

`waterStressLevel` and `minimumTurn` are **named lists** with one entry
per phenological phase:

- `vegetative`   — from transplanting to flowering,
- `reproductive` — from flowering to mid-ripening,
- `ripening`     — from mid-ripening to harvest.

Typical defaults:

```{r setup}
library(cumba)

waterStressLevel <- list(vegetative = 0.5,
                         reproductive = 0.5,
                         ripening = 0.5)

minimumTurn <- list(vegetative = 4,
                    reproductive = 4,
                    ripening = 4)
```

Lower thresholds (e.g. `0.3`) impose more deficit; higher thresholds
(e.g. `0.7`) imply more frequent watering.

# A complete walkthrough on `tomatoFoggia`

```{r run}
library(dplyr)
library(ggplot2)

data(tomatoFoggia)
weather <- tomatoFoggia$weather

out <- cumba_scenario(
  weather          = weather,
  param            = cumbaParameters,
  estimateRad      = TRUE,
  estimateET0      = TRUE,
  transplantingDOY = 120,
  waterStressLevel = waterStressLevel,
  minimumTurn      = minimumTurn,
  fullOut          = FALSE
)

head(out)
```

## Compare two irrigation strategies

Run a *full* and a *deficit* scenario and compare yield and Brix at
harvest:

```{r compare}
full_irr <- list(vegetative = 0.7, reproductive = 0.7, ripening = 0.6)
deficit  <- list(vegetative = 0.4, reproductive = 0.4, ripening = 0.3)

run_strategy <- function(level, label) {
  cumba_scenario(
    weather          = weather,
    param            = cumbaParameters,
    waterStressLevel = level,
    minimumTurn      = minimumTurn,
    transplantingDOY = 120,
    fullOut          = FALSE
  ) |>
    mutate(strategy = label)
}

scen <- bind_rows(
  run_strategy(full_irr, "Full irrigation"),
  run_strategy(deficit,  "Deficit irrigation")
)

harvest <- scen |>
  group_by(strategy, site, year) |>
  slice_tail(n = 1) |>
  ungroup() |>
  select(strategy, site, year, yield, brix)

ggplot(harvest, aes(strategy, yield, fill = strategy)) +
  geom_boxplot() +
  labs(y = "Yield (q ha-1)", x = NULL,
       title = "Yield response to irrigation strategy") +
  theme(legend.position = "none")

ggplot(harvest, aes(strategy, brix, fill = strategy)) +
  geom_boxplot() +
  labs(y = "Brix (°)", x = NULL,
       title = "Brix response to irrigation strategy") +
  theme(legend.position = "none")
```

## Daily soil-water and irrigation events

```{r daily}
ggplot(scen |> filter(year == min(year)),
       aes(doy, swc, colour = strategy)) +
  geom_line() +
  geom_point(aes(y = irrigation), size = 0.6) +
  labs(x = "DOY", y = "Soil water content (%) / Irrigation (mm)",
       title = "Soil water dynamics under two strategies")
```

# Notes

- `cumba_scenario()` does **not** require an `irrigation_df`: the
  schedule is generated internally from the stress thresholds.
- The simulation window starts at `transplantingDOY` and ends at most
  150 days later (typical processing-tomato cycle length).
- The crop coefficient (`Kc`) is computed exactly as in
  `cumba_experiment()`: it follows `kcIni` until full canopy, then
  decreases linearly during senescence based on the achieved maximum
  light interception. This guarantees that yield and Brix predictions
  are directly comparable between *experiment* and *scenario* runs.
