---
title: "Run an experiment with cumba_experiment()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Run an experiment with cumba_experiment()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE,
  eval    = FALSE
)
```

# When to use `cumba_experiment()`

Use `cumba_experiment()` when you have **observed experiments** with
their actual irrigation schedule and you want CUMBA to reproduce them.
This is the function used during **model calibration** against
yield/Brix observations.

The companion function `cumba_scenario()` is meant for *what-if*
deficit-irrigation studies, where irrigation is triggered automatically
based on phase-specific water-stress thresholds.

# Inputs

`cumba_experiment()` requires:

- a daily `weather` table with `Site`, `Tx`, `Tn`, `P`, `DATE`
  and either `Lat` (when `estimateRad = TRUE`) or `Rad`;
- a parameter object `param` (the bundled `cumbaParameters` works);
- an `irrigation_df` with `ID`, `Site`, `YEAR`, `DATE`, `WVOL`.

# A complete walkthrough on `tomatoFoggia`

```{r run}
library(cumba)
library(dplyr)
library(ggplot2)

data(tomatoFoggia)

weather       <- tomatoFoggia$weather
irrigation_df <- tomatoFoggia$irrigation
production    <- tomatoFoggia$production

out <- cumba_experiment(
  weather       = weather,
  param         = cumbaParameters,
  irrigation_df = irrigation_df,
  estimateRad   = TRUE,
  estimateET0   = TRUE,
  fullOut       = FALSE
)

head(out)
```

## Inspect simulated yield and Brix at harvest

Each (site × year × experiment) ends when the cycle is complete. The
last day of each experiment carries the harvest values:

```{r harvest}
harvest <- out |>
  group_by(site, year, experiment) |>
  slice_tail(n = 1) |>
  ungroup() |>
  select(site, year, experiment, yield, brix)

harvest
```

## Compare with observations

`tomatoFoggia$production` provides observed `yield_ref` (t ha⁻¹) and
`brix_ref` (°Brix). Yields in the short output are in q ha⁻¹, so divide
by 10:

```{r compare}
cmp <- harvest |>
  mutate(experiment = as.integer(experiment)) |>
  inner_join(production, by = c("experiment" = "ID")) |>
  mutate(yield_t = yield / 10)

ggplot(cmp, aes(yield_ref, yield_t)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Observed yield (t ha-1)",
       y = "Simulated yield (t ha-1)",
       title = "cumba_experiment - yield calibration check")

ggplot(cmp, aes(brix_ref, brix)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  labs(x = "Observed Brix (°)",
       y = "Simulated Brix (°)",
       title = "cumba_experiment - Brix calibration check")
```

## Daily dynamics with `fullOut = TRUE`

For diagnostics or calibration of internal variables, pass
`fullOut = TRUE`:

```{r full}
out_full <- cumba_experiment(
  weather       = weather,
  param         = cumbaParameters,
  irrigation_df = irrigation_df,
  fullOut       = TRUE
)

names(out_full)
```

You then have access to `kc`, `etR`, `fIntAct`, `carbonStateAct`,
`floweringStateIde`, `fruitWaterContentAct`, etc. for every day.

# Tips

- Cardinal temperatures in `param` must satisfy `Tmax > Topt > Tbase`
  and the soil must satisfy `FieldCapacity > WiltingPoint`; otherwise
  the function stops with a clear error.
- If you supply `param` as a tibble of values (instead of a list of
  lists), make sure all the required parameter names are present —
  the function checks them at start-up.
- Run-time scales linearly with the number of (site × year × ID)
  combinations: each is simulated day-by-day independently.
