---
title: "Calibration of ET0 and Radiation for Tomato Irrigation with cumba"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration of ET0 and Radiation for Tomato Irrigation with cumba}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE,
  eval    = FALSE
)
```

# Goal

This vignette demonstrates how to use **cumba** to assist tomato
farmers in irrigation scheduling by calibrating and estimating
**reference evapotranspiration (ET₀)** and **solar radiation (RAD)**
from weather-station data.

The workflow follows three progressive steps:

- **Calibration** — estimate ET₀ and RAD with the Hargreaves-style
  formulae embedded in `cumba_scenario()`/`cumba_experiment()` and
  compare them with observed values from a sub-period.
- **Parameter averaging** — derive robust mean coefficients across
  calibration sites/periods.
- **Validation** — apply the averaged parameters to independent
  datasets and compare estimated vs observed ET₀ and RAD.

We rely on open-access weather data (NASA POWER or local agromet
stations) and simple crop metadata (sowing/harvest dates for tomato
experiments).

# Packages

```{r libraries}
library(cumba)
library(dplyr)
library(ggplot2)
library(lubridate)
```

# Data

The package ships with a real-world weather dataset from Foggia
(processing-tomato area, Apulia, Italy). It already contains daily
`Tx`, `Tn`, `P`, `Rad` and the columns `Site` and `Lat` required by
the model.

```{r data}
data(tomatoFoggia)
weather <- tomatoFoggia$weather
str(weather)
```

If you want to reproduce the calibration on your own data, a CSV with
the same columns is shipped under `testFiles/weather_foggia.csv`.

# Step 1 — Run the model with internal ET₀/Rad estimation

Set `estimateRad = TRUE` and `estimateET0 = TRUE` to have CUMBA
compute radiation and ET₀ internally using temperature-based
formulae:

```{r run}
out <- cumba_scenario(
  weather     = weather,
  param       = cumbaParameters,
  estimateRad = TRUE,
  estimateET0 = TRUE,
  fullOut     = TRUE
)

out_sim <- out |>
  dplyr::select(year, doy, et0, radiation)

head(out_sim)
```

# Step 2 — Compare simulated vs observed

`tomatoFoggia$weather` provides observed daily radiation `Rad`
(MJ m⁻² d⁻¹). Joining it with the simulated outputs gives the
calibration scatter plots:

```{r compare}
obs <- weather |>
  mutate(year = year(DATE), doy = yday(DATE)) |>
  select(year, doy, Rad)

cmp <- out_sim |>
  inner_join(obs, by = c("year", "doy"))

ggplot(cmp, aes(radiation, Rad)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  xlim(0, 35) + ylim(0, 35) +
  labs(x = "Simulated radiation (MJ m-2 d-1)",
       y = "Observed radiation (MJ m-2 d-1)",
       title = "Hargreaves-based radiation vs. measured")

ggplot(cmp, aes(doy)) +
  stat_summary(geom = "ribbon", aes(y = radiation), alpha = 0.4) +
  stat_summary(geom = "ribbon", aes(y = Rad), fill = "grey50", alpha = 0.4) +
  labs(x = "DOY", y = "Radiation (MJ m-2 d-1)",
       title = "Seasonal pattern (simulated vs observed)")
```

# Step 3 — Validation on an independent year

Hold out one season and compare:

```{r validate}
cal_years <- sort(unique(cmp$year))
val_year  <- tail(cal_years, 1)

val <- cmp |> filter(year == val_year)

ggplot(val, aes(doy)) +
  geom_line(aes(y = radiation, colour = "simulated")) +
  geom_line(aes(y = Rad,       colour = "observed"))  +
  labs(x = "DOY", y = "Radiation (MJ m-2 d-1)",
       title = paste("Validation - year", val_year),
       colour = NULL)
```

# Going further

If your weather station provides observed ET₀, set
`estimateET0 = FALSE` and add an `ET0` column to `weather`. CUMBA will
then use the observed values directly, isolating the radiation
estimator from the ET₀ estimator during calibration.
