---
title: "Calibration of ET₀ and Radiation for Tomato Irrigation with cumbà"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration of ET₀ and Radiation for Tomato Irrigation with cumbà}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
css: vignette.css
---

<style>
.main-container { max-width: 1200px !important; }
table, .table, .table-condensed { font-size: 85% !important; }
caption { white-space: nowrap; }
</style>

# 🎯 Goal

This vignette demonstrates how to use cumbà to assist tomato farmers in irrigation scheduling by calibrating and estimating reference evapotranspiration (ET₀) and solar radiation (RAD) using weather station data.

The workflow follows three progressive steps:

- Calibration: estimate ET₀ and RAD parameters using a subset of weather station records.
- Parameter averaging: derive robust mean coefficients across calibration sites/periods.
- Validation: apply averaged parameters to independent datasets and compare estimated vs observed ET₀ and RAD.

We rely on open access weather data (NASA POWER or local agromet stations) and simple crop metadata (sowing/harvest dates for tomato experiments).

```{r options, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE
)
```

# 📦 Packages
We load cumbà plus helper libraries for wrangling, visualization, and reproducible tables.

```{r libraries, echo=T}
library(cumba)
library(tidyverse)
library(lubridate)
library(scales)
library(kableExtra)
```

# 📥 Data loading

We use two data sources:

- weather_data: daily records from a weather station (temperature, humidity, wind, sunshine or cloud cover).
- reference_data: observed ET₀ (from lysimeter or FAO-56 Penman–Monteith reference calculations).

# 🛰 Weather station data

Example: daily max/min temperature, humidity, wind speed, and sunshine hours for a tomato-growing region.

# 🧪 Step 1 – Calibration of ET₀ and RAD

We calibrate ET₀ and RAD estimation methods against observed reference data.

👉 Workflow:

- Select a subset of days/sites for calibration.
- Estimate ET₀ and RAD using initial formulae (e.g., Hargreaves, Angström, FAO-56 PM).
- Adjust coefficients (α, β, Krs) via optimization.
- Collect estimated parameters.

```{r simulations, echo=T}


weather_df<- read.csv("C:\\Users\\feder\\OneDrive - Università degli studi della Tuscia di Viterbo\\Desktop\\Git\\cumba_R_package\\testFiles\\weather_foggia.csv") |> 
  mutate(Rg=as.numeric(Rg),
         ET0 = as.numeric(ET0),
         date = as.Date(MDATE,format='%m/%d/%Y'),
         doy=yday(date),
         year=year(date)) |> 
  rename(Tx=TMAX,Tn=TMIN,P=RAIN,DATE=date) |> 
  mutate(Site='Foggia',Lat=41.2645)

ggplot(weather_df, aes(x=doy))+
  stat_summary(geom='ribbon',aes(y=Rg))+
  stat_summary(geom='ribbon',aes(y=ET0),fill='blue')

weather<-weather_df |> filter(year(DATE)==1951)
param<-cumba::cumbaParameters
out<-cumba_scenario(weather_df,
                    cumba::cumbaParameters, fullOut = T)

out_sim<-out |> 
  dplyr::select(year,doy,et0,radiation) |> 
  left_join(weather_df)


ggplot(out_sim,aes(x=doy))+
  stat_summary(geom='ribbon',aes(y=et0))+#simulated
  stat_summary(geom='ribbon',aes(y=ET0),fill='grey')

ggplot(out_sim,aes(x=doy))+
  stat_summary(geom='ribbon',aes(y=radiation))+#simulated
  stat_summary(geom='ribbon',aes(y=Rg),fill='grey')


ggplot(out_sim,aes(x=radiation,y=Rg))+
  geom_point()+
  ylim(0,35)+
  xlim(0,35)+
  geom_smooth(method='lm')



```